---
title: "Classificaiton Trees on Cumulative Risk"
author: "Lyndsey Umsted"
date: "2022-11-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading neccessary libraries:
```{r}
library(tidyverse)
library(tidymodels)
library(ISLR)
library(rpart.plot)
library(vip)
library(janitor)
library(randomForest)
library(xgboost)
```

Splitting into training and testing
```{r}
setwd("C:/Users/18586/Desktop/PSTAT 131/PSTAT-131-final-project/models")
load("pandemic_cum.rda")

set.seed(2002)

pandemic_cum <- pandemic_cum %>%
  dplyr::select(-c("prop", "confirmed_cases", "population", "ID_co"))

pandemic_cum_split <- initial_split(pandemic_cum, prop = 0.80, strata = risk)

pandemic_cum_train <- training(pandemic_cum_split)
pandemic_cum_test <- testing(pandemic_cum_split)
```

K-fold Cross Validation:
```{r}
pandemic_folds <- vfold_cv(pandemic_cum_train, v = 10, strata = risk)  # 10-fold CV
```

Fitting classification tree:
```{r}
tree_spec <- decision_tree() %>%
  set_engine("rpart")

class_tree_spec <- tree_spec %>%
  set_mode("classification")
```


Fit the model to the training data:
```{r}
class_tree_fit <- class_tree_spec %>%
  fit(risk ~., data = pandemic_cum_train)
```



Visualize the decision tree:
```{r}
class_tree_fit %>%
  extract_fit_engine() %>%
  rpart.plot()
```

Training set accuracy:
```{r}
augment(class_tree_fit, new_data = pandemic_cum_train) %>%
  accuracy(truth = risk, estimate = .pred_class)
```

Confusion Matrix:
```{r}
augment(class_tree_fit, new_data = pandemic_cum_train) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

Accuracy on testing set:
```{r}
augment(class_tree_fit, new_data = pandemic_cum_test) %>%
  accuracy(truth = risk, estimate = .pred_class)
```


Confusion Matrix:
```{r}
augment(class_tree_fit, new_data = pandemic_cum_test) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

Let us try to tune the cost_complexity of the decision tree, or the pruning penalty, to find a more optimal complexity. We use the class_tree_spec object and use the set_args() function to specify that we want to tune cost_complexity. This is then passed directly into the workflow object to avoid creating an intermediate object.

```{r}
class_tree_wf <- workflow() %>%
  add_model(class_tree_spec %>% set_args(cost_complexity = tune())) %>%
  add_formula(risk ~ .)
```

To be able to tune the hyperparameter, we need 2 more objects – a resamples object (we will use a k-fold cross-validation data set), and a grid of values to try. Since we are only tuning one hyperparameter, a regular grid is probably simplest.

```{r}
set.seed(3435)
pandemic_fold <- vfold_cv(pandemic_cum_train, v = 10, strata = risk)

param_grid <- grid_regular(cost_complexity(range = c(-3, 0)), levels = 10)

tune_res_st <- tune_grid(
  class_tree_wf, 
  resamples = pandemic_fold, 
  grid = param_grid, 
  metrics = metric_set(roc_auc)
)

```

Using autoplot() shows which values of cost_complexity appear to produce the highest roc_auc:
```{r}
autoplot(tune_res_st)
```

We can now select the best performing value with select_best(), finalize the workflow by updating the value of cost_complexity, and fit the model on the full training data set.

```{r}
best_complexity <- select_best(tune_res_st)

class_tree_final <- finalize_workflow(class_tree_wf, best_complexity)

class_tree_final_fit <- fit(class_tree_final, data = pandemic_cum_train)
```

At last we can visualize the model, and we see that the better-performing model is much less complex than the original model we fit.

```{r}
class_tree_final_fit %>%
  extract_fit_engine() %>%
  rpart.plot()
```

Accuracy on testing set:
```{r}
class_tree_acc <- augment(class_tree_final_fit, new_data = pandemic_cum_test) %>%
  accuracy(truth = risk, estimate = .pred_class)

class_tree_acc
```


Confusion Matrix:
```{r}
augment(class_tree_final_fit, new_data = pandemic_cum_test) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

ROC Curve:
```{r}
roc <- augment(class_tree_final_fit, pandemic_cum_test)

roc %>%
  roc_curve(risk, c(.pred_low, .pred_moderate, .pred_high)) %>%
  autoplot()
```

AUC:
```{r}
single_tree_roc <- roc %>%
  roc_auc(risk, c(.pred_low, .pred_moderate, .pred_high))
single_tree_roc
```

Bagging and Random Forest:

```{r}
bagging_spec <- rand_forest(mtry = .cols()) %>%
  set_engine("randomForest", importance = TRUE) %>%
  set_mode("classification")
```

fit the model:
```{r}
bagging_fit <- fit(bagging_spec, risk ~ ., 
                   data = pandemic_cum_train)
```

testing performance:
```{r}
bagged_acc <- augment(bagging_fit, new_data = pandemic_cum_test) %>%
  accuracy(truth = risk, estimate = .pred_class)
bagged_acc
```

Confusion Matrix:
```{r}
augment(bagging_fit, new_data = pandemic_cum_test) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

ROC Curve:
```{r}
roc <- augment(bagging_fit, pandemic_cum_test)

roc %>%
  roc_curve(risk, c(.pred_low, .pred_moderate, .pred_high)) %>%
  autoplot()
```

AUC:
```{r}
bagging_roc <- roc %>%
  roc_auc(risk, c(.pred_low, .pred_moderate, .pred_high))
bagging_roc
```



Variable Importance:
```{r}
vip(bagging_fit)
```

looking at the random forest:
```{r}
rf_spec <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

tune_wf <- workflow() %>%
  add_formula(risk ~ Mean_Tmax + Urban + PovertyRate + MedianFamilyIncome + Traffic + Solid.Waste + Asthma + Low.Birth.Weight + Cardiovascular.Disease + Education + Linguistic.Isolation + Unemployment + under_10_. + Age11_to_64_. + over_65_. + Hispanic_. + White_. + African_Am_. + Asian_Am_. + Native_Am_. + Other_ethnicity_.) %>%
  add_model(rf_spec)

set.seed(2002)
pandemic_fold <- vfold_cv(pandemic_cum_train, v = 10, strata = risk)
```

Creating a Regular Grid:
```{r}
rf_grid <- grid_regular(mtry(range = c(2, 20)), trees(range = c(6, 10)), min_n(range = c(3, 5)), levels = 10)
```


Tuning the model:
```{r, eval=FALSE}
tune_res_rf <- tune_grid(
  tune_wf,
  resamples = pandemic_fold,
  grid = rf_grid, 
  metrics = metric_set(roc_auc)
)


save(tune_res_rf, file = "tune_res_rf.rda")
```


```{r}
load("tune_res_rf.rda")
tune_res_rf

autoplot(tune_res_rf)
```

Selecting best combination of mtry() trees() and min_n()

```{r}
best <- select_best(tune_res_rf)

rf_final <- finalize_workflow(tune_wf, best)

rf_final_fit <- fit(rf_final, data = pandemic_cum_train)
```



accuracy on training:
```{r}
augment(rf_final_fit, new_data = pandemic_cum_train) %>%
  accuracy(truth = risk, estimate = .pred_class)
```


accuracy on testing:
```{r}
rf_acc <- augment(rf_final_fit, new_data = pandemic_cum_test) %>%
  accuracy(truth = risk, estimate = .pred_class)
rf_acc
```


Confusion Matrix:
```{r}
augment(rf_final_fit, new_data = pandemic_cum_test) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

ROC Curve:
```{r}
roc <- augment(rf_final_fit, pandemic_cum_test)

roc %>%
  roc_curve(risk, c(.pred_low, .pred_moderate, .pred_high)) %>%
  autoplot()
```

AUC:
```{r}
rf_roc <- roc %>%
  roc_auc(risk, c(.pred_low, .pred_moderate, .pred_high))
rf_roc
```

Variable importance:
```{r}
vip(rf_fit)
```


Comparing Model Performance:

Accuracy:
```{r}
performance <- c(class_tree_acc$.estimate, bagging_roc$.estimate, rf_acc$.estimate)
models <- c("Single Decision Tree", "Bagged Tree", "Random Forest")
results <- tibble(Accuracy = performance, models = models)
results %>% 
  arrange(-performance)
```


AUC:
```{r}
performance <- c(single_tree_roc$.estimate, bagged_acc$.estimate, rf_roc$.estimate)
models <- c("Single Decision Tree", "Bagged Tree", "Random Forest")
results <- tibble(AUC = performance, models = models)
results %>% 
  arrange(-performance)
```

