---
title: "Classificaiton Trees on Cumulative Risk"
author: "Lyndsey Umsted"
date: "2022-11-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading neccessary libraries:
```{r}
library(tidyverse)
library(tidymodels)
library(ISLR)
library(rpart.plot)
library(vip)
library(janitor)
library(randomForest)
library(xgboost)
```

Splitting into training and testing
```{r}
setwd("C:/Users/18586/Desktop/PSTAT 131/PSTAT-131-final-project/models")
load("pandemic_cum.rda")

set.seed(2002)

pandemic_cum <- pandemic_cum %>%
  select(-c("prop", "confirmed_cases", "population", "ID_co", "proplog"))

pandemic_cum_split <- initial_split(pandemic_cum, prop = 0.80, strata = risk)

pandemic_cum_train <- training(pandemic_cum_split)
pandemic_cum_test <- testing(pandemic_cum_split)
```

Fitting classification tree:
```{r}
tree_spec <- decision_tree() %>%
  set_engine("rpart")

class_tree_spec <- tree_spec %>%
  set_mode("classification")
```


Fit the model:
```{r}
class_tree_fit <- class_tree_spec %>%
  fit(risk ~., data = pandemic_cum_train)
```


Visualize the decision tree:
```{r}
class_tree_fit %>%
  extract_fit_engine() %>%
  rpart.plot()
```

Training set accuracy:
```{r}
augment(class_tree_fit, new_data = pandemic_cum_train) %>%
  accuracy(truth = risk, estimate = .pred_class)
```

Confusion Matrix:
```{r}
augment(class_tree_fit, new_data = pandemic_cum_train) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

Accuracy on testing set:
```{r}
augment(class_tree_fit, new_data = pandemic_cum_test) %>%
  accuracy(truth = risk, estimate = .pred_class)
```


Confusion Matrix:
```{r}
augment(class_tree_fit, new_data = pandemic_cum_test) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

Let us try to tune the cost_complexity of the decision tree, or the pruning penalty, to find a more optimal complexity. We use the class_tree_spec object and use the set_args() function to specify that we want to tune cost_complexity. This is then passed directly into the workflow object to avoid creating an intermediate object.

```{r}
class_tree_wf <- workflow() %>%
  add_model(class_tree_spec %>% set_args(cost_complexity = tune())) %>%
  add_formula(risk ~ .)
```

To be able to tune the hyperparameter, we need 2 more objects – a resamples object (we will use a k-fold cross-validation data set), and a grid of values to try. Since we are only tuning one hyperparameter, a regular grid is probably simplest.

```{r}
set.seed(3435)
pandemic_fold <- vfold_cv(pandemic_cum_train)

param_grid <- grid_regular(cost_complexity(range = c(-3, -1)), levels = 10)

tune_res <- tune_grid(
  class_tree_wf, 
  resamples = pandemic_fold, 
  grid = param_grid, 
  metrics = metric_set(accuracy)
)

```

Using autoplot() shows which values of cost_complexity appear to produce the highest accuracy:
```{r}
autoplot(tune_res)
```

We can now select the best performing value with select_best(), finalize the workflow by updating the value of cost_complexity, and fit the model on the full training data set.

```{r}
best_complexity <- select_best(tune_res)

class_tree_final <- finalize_workflow(class_tree_wf, best_complexity)

class_tree_final_fit <- fit(class_tree_final, data = pandemic_cum_train)
```

At last we can visualize the model, and we see that the better-performing model is much less complex than the original model we fit.

```{r}
class_tree_final_fit %>%
  extract_fit_engine() %>%
  rpart.plot()
```

Accuracy on testing set:
```{r}
augment(class_tree_final_fit, new_data = pandemic_cum_test) %>%
  accuracy(truth = risk, estimate = .pred_class)
```


Confusion Matrix:
```{r}
augment(class_tree_final_fit, new_data = pandemic_cum_test) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

ROC Curve:
```{r}
roc <- augment(class_tree_final_fit, pandemic_cum_test)

roc %>%
  roc_curve(risk, c(.pred_low, .pred_moderate, .pred_high)) %>%
  autoplot()
```

AUC:
```{r}
roc %>%
  roc_auc(risk, c(.pred_low, .pred_moderate, .pred_high))
```







