---
title: "Lasso Regression"
author: "Lyndsey Umsted"
date: "2022-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Libraries:
```{r}
library(MASS)
library(tidyverse)
library(tidymodels)
library(ISLR)
library(ISLR2)
library(glmnet)
tidymodels_prefer()
library(discrim)
library(dplyr)
```


Splitting into training and testing
```{r}
setwd("C:/Users/18586/Desktop/PSTAT 131/PSTAT-131-final-project/models")
load("pandemic_cum.rda")

set.seed(2002)

pandemic_cum <- pandemic_cum %>%
  select(-c("prop", "confirmed_cases", "population"))

pandemic_cum_split <- initial_split(pandemic_cum, prop = 0.80, strata = risk)

pandemic_cum_train <- training(pandemic_cum_split)
pandemic_cum_test <- testing(pandemic_cum_split)



pandemic_folds <- vfold_cv(pandemic_cum_train[,-1], v = 10, strata = risk)  # 10-fold CV

```

Creating Recipe:
```{r}
pandemic_cum_rec <- recipe(risk ~ ., data = pandemic_cum_train[,-1]) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_predictors()) 
```

We'll be fitting and tuning an elastic net, tuning `penalty` and `mixture` (use `multinom_reg` with the `glmnet` engine).

Set up this model and workflow. Create a regular grid for `penalty` and `mixture` with 10 levels each; `mixture` should range from 0 to 1. For this assignment, we'll let `penalty` range from -5 to 5 (it's log-scaled).

```{r}
elastic_spec <- 
  multinom_reg(penalty = tune(), mixture = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("glmnet")

elastic_workflow <- workflow() %>% 
  add_recipe(pandemic_cum_rec) %>% 
  add_model(elastic_spec)

penalty_grid <- grid_regular(penalty(range = c(-6, 6)), mixture(range = c(0,1)), levels = 15)

penalty_grid
```


Fit the models to your folded data using `tune_grid()`.

```{r, message = FALSE}
tune_res_elastic <- tune_grid(
  elastic_workflow,
  resamples = pandemic_folds, 
  grid = penalty_grid
)

tune_res_elastic

elastic_plot <- autoplot(tune_res_elastic)
elastic_plot
save(elastic_plot, file = "elastic_plot.rda")
```

Best penalty accuracy on testing set:
```{r}
best_penalty <- select_best(tune_res_elastic, metric = "roc_auc")
best_penalty


elastic_final <- finalize_workflow(elastic_workflow, best_penalty)

elastic_final_fit <- fit(elastic_final, data = pandemic_cum_train[,-1])

elastic_acc <- augment(elastic_final_fit, new_data = pandemic_cum_test[,-1]) %>%
  accuracy(truth = risk, estimate = .pred_class)

elastic_acc
```

AUC:
```{r}
roc <- augment(elastic_final_fit, pandemic_cum_test[,-1])

elastic_auc <- roc %>%
  roc_auc(risk, c(.pred_low, .pred_moderate, .pred_high))

elastic_auc
```



Then create plots of the different ROC curves, one per level of the outcome. 

```{r}
elastic_roc <- roc %>%
  roc_curve(risk, c(.pred_low, .pred_moderate, .pred_high)) %>%
  autoplot()

elastic_roc
```

Also make a heat map of the confusion matrix.

confusion matrix:
```{r}
augment(elastic_final_fit, new_data = pandemic_cum_test[,-1]) %>%
  conf_mat(truth = risk, estimate = .pred_class) %>%
  autoplot(type = "heatmap")
```

Saving Performance:
```{r}
save(elastic_acc, file = "elastic_acc.rda")
save(elastic_roc, file = "elastic_roc.rda")
save(elastic_auc, file = "elastic_auc.rda")
```


