---
title: "KNN Model for Cumulative Data"
author: "Lyndsey Umsted"
date: "2022-10-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading necessary libraries
```{r}
#install.packages("kknn")
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(corrplot)
library(ggthemes)
tidymodels_prefer()
library(kknn)
```

Splitting into training and testing
```{r}
setwd("C:/Users/18586/Desktop/PSTAT 131/PSTAT-131-final-project/models")
load("pandemic_cum.rda")

set.seed(2002)

pandemic_cum <- pandemic_cum %>%
  select(-c("prop", "confirmed_cases", "population"))

pandemic_cum_split <- initial_split(pandemic_cum, prop = 0.80, strata = risk)

pandemic_cum_train <- training(pandemic_cum_split)
pandemic_cum_test <- testing(pandemic_cum_split)
```

Creating Recipe:
```{r}
pandemic_cum_rec <- recipe(risk ~ ., data = pandemic_cum_train[,-1]) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_predictors())
```

K-fold Cross Validation:
```{r}
pandemic_folds <- vfold_cv(pandemic_cum_train[,-1], v = 10, strata = risk)  # 10-fold CV
```

KNN Model:
```{r}
knn_model <- nearest_neighbor(
  mode = "classification",
  engine = "kknn", 
  neighbors = 10,
  weight_func = "rectangular", 
  dist_power = NULL
)
```

Workflow:
```{r}
knn_wflow_cum <- workflow() %>% 
  add_model(knn_model) %>% 
  add_recipe(pandemic_cum_rec)
```

Fit KNN model to folded data:
```{r}
knn_fit_cum_folds <- fit_resamples(knn_wflow_cum, pandemic_folds)
save(knn_fit_cum_folds, file = "knn_fit_cum_folds.rda")
```

Collect Metrics
```{r}
collect_metrics(knn_fit_cum_folds)
```


Fit model to training data:
```{r}
knn_fit_cum <- fit(knn_wflow_cum, pandemic_cum_train[,-1])
```


Generating predicted values for each observation of the training set:
```{r}
pandemic_cum_knn_pred <- predict(knn_fit_cum, new_data = pandemic_cum_train[,-1] %>% select(-risk))
pandemic_cum_knn_pred %>% 
  head()
```

Attaching a column of the actual observed confirmed cases:
```{r}
pandemic_cum_knn_pred <- bind_cols(pandemic_cum_knn_pred, pandemic_cum_train[,-1] %>% select(risk))
pandemic_cum_knn_pred %>% 
  head()
```

```{r}
augment(knn_fit_cum, new_data = pandemic_cum_train[,-1]) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

Accuracy on Training

```{r}
knn_acc <- augment(knn_fit_cum, new_data = pandemic_cum_train[,-1]) %>%
  accuracy(truth = risk, estimate = .pred_class)
knn_acc
```

Testing Data:

```{r}
pandemic_cum_knn_pred <- predict(knn_fit_cum, new_data = pandemic_cum_test[,-1] %>% select(-risk))
pandemic_cum_knn_pred %>% 
  head()
```

Attaching a column of the actual observed confirmed cases:
```{r}
pandemic_cum_knn_pred <- bind_cols(pandemic_cum_knn_pred, pandemic_cum_test[,-1] %>% select(risk))
pandemic_cum_knn_pred %>% 
  head()
```

```{r}
augment(knn_fit_cum, new_data = pandemic_cum_test[,-1]) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```

```{r}
knn_acc_test <- augment(knn_fit_cum, new_data = pandemic_cum_test[,-1]) %>%
  accuracy(truth = risk, estimate = .pred_class)
knn_acc_test
```


ROC Curve:
```{r}
roc <- augment(knn_fit_cum, pandemic_cum_train[,-1])

knn_roc <- roc %>%
  roc_curve(risk, c(.pred_low, .pred_moderate, .pred_high)) %>%
  autoplot()

knn_roc
```


AUC:
```{r}
knn_auc <- roc %>%
  roc_auc(risk, c(.pred_low, .pred_moderate, .pred_high))
knn_auc
```

Saving Performance:
```{r}
save(knn_acc, file = "knn_acc.rda")
save(knn_auc, file = "knn_auc.rda")
save(knn_roc, file = "knn_roc.rda")
```


