---
title: "Linear Regression Model on Weekly Data"
author: "Lyndsey Umsted"
date: "2022-10-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I will be fitting a linear regression model to the Weekly Data

Loading neccessary libraries
```{r}
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(corrplot)
library(ggthemes)
tidymodels_prefer()
```

Splitting into training and testing
```{r}
load("pandemic_weekly.rda")
set.seed(24)

pandemic_weekly_split <- initial_split(pandemic_weekly, prop = 0.80, strata = new_cases)

pandemic_weekly_train <- training(pandemic_weekly_split)
pandemic_weekly_test <- testing(pandemic_weekly_split)
```


Creating Recipe:
```{r}
pandemic_weekly_rec <- recipe(new_cases ~ ., data = pandemic_weekly_train) %>%
  step_dummy(all_nominal_predictors())
```


Linear Regression Model:
```{r}
lm_model <- linear_reg() %>% 
  set_engine("lm")
```

Workflow:
```{r}
lm_wflow_weekly <- workflow() %>% 
  add_model(lm_model) %>% 
  add_recipe(pandemic_weekly_rec)
```

Fit linear model to training data:
```{r}
lm_fit_weekly <- fit(lm_wflow_weekly, pandemic_weekly_train)
```

We can view the model results:

```{r}
lm_fit_weekly %>% 
  # This returns the parsnip object:
  extract_fit_parsnip() %>% 
  # Now tidy the linear model object:
  tidy()
```

Generating preicted values for each observation of the training set:
```{r}
pandemic_weekly_train_res <- predict(lm_fit_weekly, new_data = pandemic_weekly_train %>% select(-new_cases))
pandemic_weekly_train_res %>% 
  head()
```


Attaching a column of the actual observed confirmed cases:
```{r}
pandemic_weekly_train_res <- bind_cols(pandemic_weekly_train_res, pandemic_weekly_train %>% select(new_cases))
pandemic_weekly_train_res %>% 
  head()
```


Plot of predicted values versus actual values of confirmed cases:
```{r}
pandemic_weekly_train_res %>% 
  ggplot(aes(x = .pred, y = new_cases)) +
  geom_point(alpha = 0.2) +
  geom_abline(lty = 2) + 
  theme_bw() +
  coord_obs_pred()
```

Looking at RMSE on the training data:
```{r}
rmse(pandemic_weekly_train_res, truth = new_cases, estimate = .pred)
```

View metric set:
```{r}
pandemic_weekly_metrics <- metric_set(rmse, rsq, mae)
pandemic_weekly_metrics(pandemic_weekly_train_res, truth = new_cases, 
                estimate = .pred)
```


```{r}
simple_ln_fit_weekly <- lm(new_cases ~ ., pandemic_weekly_train)
summary(simple_ln_fit_weekly)
```