---
title: "Polynomial Regression Model on Weekly Data"
author: "Lyndsey Umsted"
date: "2022-10-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I will be fitting a Polynomial Regression Model to the Weekly Data

Loading necessary libraries
```{r}
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(corrplot)
library(ggthemes)
tidymodels_prefer()
```

Splitting into training and testing
```{r}
set.seed(24)

pandemic_weekly_split <- initial_split(pandemic_weekly, prop = 0.80, strata = new_cases)

pandemic_weekly_train <- training(pandemic_weekly_split)
pandemic_weekly_test <- testing(pandemic_weekly_split)
```

First, we set up a linear regression specification:

```{r}
lm_spec <- linear_reg() %>%
  set_mode("regression") %>%
  set_engine("lm")
```

We can use the linear model specification `lm_spec` to add a preprocessing unit with `recipe()` and `step_poly()` to create the polynomial expansion of the predictors. We can combine these two with `workflow()` to create a workflow object.

```{r}
weekly_poly_rec <- recipe(new_cases ~ ., data = pandemic_weekly_train) %>%
  step_poly(all_predictors(), degree = 2)

weekly_poly_wf <- workflow() %>%
  add_recipe(weekly_poly_rec) %>%
  add_model(lm_spec)

weekly_poly_wf
```

Now fitting model to training data set:

```{r}
weekly_poly_fit <- fit(weekly_poly_wf, data = pandemic_weekly_train)
```

The testing RMSE is then calculated as:

```{r}
augment(weekly_poly_fit, new_data = pandemic_weekly_test) %>%
  rmse(truth = new_cases, estimate = .pred)
```

Above I used the degree 2 in my polynomial fit, in order to find the degree that yields the "closest fit" I need to hyperparamter tune. Using k-Fold Cross Validiation to do this

need:
- A parsnip/workflow object with one or more arguments marked for tuning,

- A `vfold_cv` rsample object of the cross-validation resamples, &

- A tibble denoting the values of hyperparameter values to be explored.


Creating a new recipe with `degree = tune()` indicates that I intend for degree to be tuned.


```{r}
weekly_poly_tuned_rec <- recipe(new_cases ~ ., data = pandemic_weekly_train) %>%
  step_poly(all_predictors(), degree = tune())

weekly_poly_tuned_wf <- workflow() %>%
  add_recipe(weekly_poly_tuned_rec) %>%
  add_model(lm_spec)
```


The next thing we need to create is the k-Fold data set. This can be done using the `vfold_cv()` function. Note that the function uses `v` instead of `k`, which is the terminology of ISLR. Common choices for k/v are 5 or 10. Here, we use 10 folds.

```{r}
pandemic_weekly_folds <- vfold_cv(pandemic_weekly_train, v = 10)
pandemic_weekly_folds
```


The result is a tibble of `vfold_splits` which is quite similar to the `rsplit` object we saw earlier.

The last thing we need is a tibble of possible values we want to explore. Each of the tuneable parameters in `tidymodels` has an associated function in the `dials` package. We need to use the `degree()` function here, and we extend the range to have a max of 10. This `dials` function is then passed to `grid_regular()` to create a regular grid of values.


```{r}
weekly_degree_grid <- grid_regular(degree(range = c(1, 10)), levels = 10)
weekly_degree_grid
```


Now that all the necessary objects have been created, we can pass them to `tune_grid()`, which will fit the models within each fold for each value specified in `degree_grid`.

```{r,include=FALSE}
weekly_tune_res <- tune_grid(
  object = weekly_poly_tuned_wf, 
  resamples = pandemic_weekly_folds, 
  grid = weekly_degree_grid,
  control = control_grid(verbose = TRUE)
)
```

Visual representation of hyperparameter performance

```{r}
autoplot(weekly_tune_res)
```


```{r}
collect_metrics(weekly_tune_res)
```


show the best performing models:

```{r}
show_best(weekly_tune_res, metric = "rmse")
```


```{r}
select_by_one_std_err(weekly_tune_res, degree, metric = "rmse")
```

```{r}
best_degree <- select_by_one_std_err(weekly_tune_res, degree, metric = "rmse")
```


Final wf with best degree:

```{r}
final_wf_weekly <- finalize_workflow(weekly_poly_wf, best_degree)

final_wf_weekly
```


```{r}
weekly_final_fit <- fit(final_wf_weekly, pandemic_weekly_train)

weekly_final_fit
```


Finally, we assess its accuracy on the testing set:

```{r}
augment(weekly_final_fit, new_data = pandemic_weekly_test) %>%
  rmse(truth = new_cases, estimate = .pred)
```

