---
title: "Random Forest Model for Covid-19 Risk"
author: "Lyndsey Umsted"
date: "2022-11-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading necessary libraries
```{r}
library(tidymodels)
library(ISLR) # For the Smarket data set
library(ISLR2) # For the Bikeshare data set
library(discrim)
library(poissonreg)
library(corrr)
library(klaR) # for naive bayes
tidymodels_prefer()
library(randomForest)
```

Splitting into training and testing
```{r}
setwd("C:/Users/18586/Desktop/PSTAT 131/PSTAT-131-final-project/models")
load("pandemic_cum.rda")

set.seed(2002)

pandemic_cum <- pandemic_cum %>%
  select(-c("prop", "confirmed_cases", "population", "ID_co", "proplog"))

pandemic_cum_split <- initial_split(pandemic_cum, prop = 0.70, strata = risk)

pandemic_cum_train <- training(pandemic_cum_split)
pandemic_cum_test <- testing(pandemic_cum_split)
```


Creating Folds
```{r}
folds <- vfold_cv(pandemic_cum_train, v = 10, repeats = 2)
```

Processing the Data

creating the recipe and assigning the steps for processing the data
```{r}
rec <- recipe(risk ~., 
  data = pandemic_cum_train) %>%   # Formula
  step_dummy(all_nominal())        # normalize numeric data to have a standard deviation of one.
  # %>%step_downsample(Target)                    # all classes should have the same frequency as the minority 
                                                     # class(not needed in our case)
```

Training the recipe data
```{r}
trained_rec <-  prep(rec, training = pandemic_cum_train, retain = TRUE)
# create the train and test set 
train_data <- as.data.frame(juice(trained_rec))
test_data  <- as.data.frame( bake(trained_rec, new_data = pandemic_cum_test))
```

The Model

We will use the {parsnip} function rand_forest() to create a random forest model and add the r-package “ranger” as the computational engine.
```{r}
#setwd("C:/Users/18586/Desktop/PSTAT 131/PSTAT-131-final-project")
# Build the model (generate the specifications of the model) 
model_spec_default <- rand_forest(mode = "classification")%>%set_engine("ranger", verbose = TRUE)
model_spec_default <- randomForest(pandemic_cum_train$risk ~ .,pandemic_cum_train)
save(model_spec_default, file = "visuals/var_imp_plot.rda")
plot(model_spec_default)
var_imp_plot <- varImpPlot(model_spec_default, cex = 1, pch = 16, main = "Variable Importance Under Random Forest Model")
```

Fit the model on the training data (train_data prepared above)
```{r}
library(tictoc)
tic()
# fit the model
model_fit_default <- model_spec_default %>%
  fit(risk ~ .,pandemic_cum_train)
toc()
## 27.85 sec elapsed

# show the configuration of the fitted model
model_fit_default
```

Predict on the testing data (test_data) and extract the model performance. How does this model perform against the holdout data (test_data, not seen before)?
```{r}
multi_metric <- metric_set(accuracy, sensitivity, specificity)

augment(model_fit_default, new_data = pandemic_cum_test) %>%
  multi_metric(truth = risk, estimate = .pred_class)
```

```{r}
augment(model_fit_default, new_data = pandemic_cum_test) %>%
  conf_mat(truth = risk, estimate = .pred_class) 
```

ROC Curve:
```{r}
roc <- augment(model_fit_default, pandemic_cum_test)

roc %>%
  roc_curve(risk, c(.pred_low, .pred_moderate, .pred_high)) %>%
  autoplot()
```


AUC:
```{r}
roc %>%
  roc_auc(risk, c(.pred_low, .pred_moderate, .pred_high))
```

