---
title: "Support Vector Machines Model"
author: "Lyndsey Umsted"
date: "2022-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading Packages:
```{r}
library(tidyverse)
library(tidymodels)
library(ISLR)
library(kernlab)
```

Data:

```{r}
setwd("C:/Users/18586/Desktop/PSTAT 131/PSTAT-131-final-project/models")
load("pandemic_cum.rda")

set.seed(2002)

pandemic_cum <- pandemic_cum %>%
  select(-c("prop", "confirmed_cases", "population", "ID_co", "proplog"))

pandemic_cum_split <- initial_split(pandemic_cum, prop = 0.80, strata = risk)

pandemic_cum_train <- training(pandemic_cum_split)
pandemic_cum_test <- testing(pandemic_cum_split)
```


Support Vector Classifier:
```{r}
svm_linear_spec <- svm_poly(degree = 1) %>%
  set_mode("classification") %>%
  set_engine("kernlab", scaled = FALSE)
```

Taking the specification, we can add a specific cost of 10 before fitting the model to the data. Using set_args() allows us to set the cost argument without modifying the model specification.


```{r}
svm_linear_fit <- svm_linear_spec %>% 
  set_args(cost = 10) %>%
  fit(risk ~ ., data = pandemic_cum_train)

svm_linear_fit
```

The kernlab models can be visualized using the plot() function if you load the kernlab package:


```{r}
svm_linear_fit <- svm_linear_spec %>% 
  set_args(cost = 0.1) %>%
  fit(risk ~ ., data = pandemic_cum_train)

svm_linear_fit
```

Let us set up a tune_grid() section to find the value of cost that leads to the highest accuracy for the SVM model.

```{r, eval = FALSE}
svm_linear_wf <- workflow() %>%
  add_model(svm_linear_spec %>% set_args(cost = tune())) %>%
  add_formula(risk ~ .)

set.seed(3435)
pandemic_cum_fold <- vfold_cv(pandemic_cum_train, strata = risk)

param_grid <- grid_regular(cost(), levels = 10)

tune_res <- tune_grid(
  svm_linear_wf, 
  resamples = pandemic_cum_fold, 
  grid = param_grid
)

save(tune_res, file = "tune_res.rda")

autoplot(tune_res)
```


Using the tune_res object and select_best() function allows us to find the value of cost that gives the best cross-validated accuracy. Finalize the workflow with finalize_workflow() and fit the new workflow on the data set.

```{r}
load("tune_res.rda")
tune_res

best_cost <- select_best(tune_res, metric = "accuracy")

svm_linear_final <- finalize_workflow(svm_linear_wf, best_cost)

svm_linear_fit <- svm_linear_final %>% fit(pandemic_cum_train)
```

Accessing the model on this testing data set shows us that the model still performs well.

```{r}
augment(svm_linear_fit, new_data = pandemic_cum_test) %>%
  conf_mat(truth = risk, estimate = .pred_class)
```






